A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Finished editing Analysis, improved the presentation of some figures, added alternate models for reference, for sensitivity analysis. Next step: Results and discussion, and conclusions.

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A spatial analysis of the environmental correlates of COVID-19 incidence in Spain},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{A spatial analysis of the environmental correlates of COVID-19 incidence
in Spain}
    \author[McMaster University]{Antonio Paez\corref{1}}
   \ead{paezha@mcmaster.ca} 
    \author[Universidad Politecnica de Cartagena]{Fernando A. Lopez}
   \ead{fernando.lopez@upct.es} 
    \author[Departamento de Economia]{Tatiane Menezes}
   \ead{tatiane.menezes@ufpe.br} 
    \author[Nucleo de Pesquisa]{Renata Cavalcanti}
   \ead{renata.vcsantos@gmail.com} 
    \author[Nucleo de Pesquisa]{Maira Galdino da Rocha Pitta}
   \ead{mgrpitta@ufpe.br} 
      \address[McMaster University]{School of Geography and Earth Sciences, McMaster University, 1281 Main
St W, Hamilton, ON, L8S 4K1, Canada}
    \address[Universidad Politecnica de Cartagena]{Departamento de Metodos Cuantitativos, Ciencias Juridicas, y Lenguas
Modernas, Universidad Politecnica de Cartagena, Calle Real Numero 3,
30201, Cartagena, Murcia, Spain}
    \address[Departamento de Economia]{Departamento de Economia, Universidade Federal de Pernambuco, Av dos
Economistas, s/n - Cidade Universitária, Recife - PE, 50670-901, Brasil}
    \address[Nucleo de Pesquisa]{Núcleo de Pesquisa em Inovação Terapêutica NUPIT / UFPE, Av.
Prof.~Moraes Rego, 1235 - Cidade Universitária, Recife, PE, CEP
50670-901, Brazil}
      \cortext[1]{Corresponding Author}

@@

\hypertarget{context-data-and-methods}{%
\section{Context, Data, and Methods}\label{context-data-and-methods}}

@@

The first variable that we consider is GDP per capita. Much has been
said about globalization and the spread of infectious
disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p. A18)}.
The growth in global connections has presented a challenge to spatial
approaches in the initial stages of disease management, when its cause
may still be unclear (Zhou and Coleman, 2016). In reference to the
earlier outbreak of SARS, van Wagner (Van Wagner, 2008) remarks how
Toronto's status as a global city was a vulnerability. In our case, we
think of GDP per capita as a marker of a region's relative position in a
network of global cities, and their potential to be further ahead in the
trajectory of the pandemic. Furthermore, wealthier regions also tend to
concentrate more activities that produce non-traded goods, including
building and construction (Hallet, 2002). Therefore, it is possible that
wealthier regions remained relatively more active even during the
lockdown. On the other hand, it is also possible the less wealthy
regions have a higher proportion of workers in manual occupations that
cannot telework.

The percentaje of older adults (over 65) is the second variable that we
consider as a control. Early evidence suggest that the case rate
mortality is higher at older ages (e.g.~The Novel Coronavirus Pneumonia
Emergency Response Epidemiology Team, 2020). It is not clear that
presence of older adults necessarily translates into higher transmission
in the population. The public health tool of choice in containing the
spread of the disease has been social distancing. In this respect, the
evidence from the field of transportation is that older adults tend to
travel less frequently, for shorter distances, and have higher rates of
immobility (e.g., Roorda et al., 2010; Morency et al., 2011; Sikder and
Pinjari, 2012). In other words, older adults are, by preference or
necessity, already in a form of social isolation. Social distancing
during the pandemic may actually reinforce that condition for older
adults, as the analysis of age-structured social contact in India,
China, and Italy of Sing and Adhikari (2020) suggests. Accordingly, our
expectation is that in the provinces with higher proportions of older
adults will tend to have similarl levels of lower social contact,
particularly since age-structured social contact matrix in Spain is
similar to Italy (Prem et al., 2017).

@@

The last control variable is the presence of mass transit systems in a
province. Every province in Spain offers some form of public
transportation, however only five provinces have higher order systems of
mass mobility, namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia.
Public transportation has been hypothesized to relate to the spread of
contagious disease by some researchers using agent-based approaches
(Perez and Dragicevic, 2009; Wang et al., 2010), and while we find scant
evidence of a link in the literature, the idea is intuitively appealing.
After all, unlike the isolation that a car offers to travellers, most
mass transit system are cauldrons of social contact.

@@

\hypertarget{methods-spatial-sur}{%
\subsection{Methods: Spatial SUR}\label{methods-spatial-sur}}

@@

The SUR-SLM model can be estimated using maximum likelihood (López et
al. (2014)) or instrumental variables (Mínguez et al. (2019)). Another
alternative methodologies could be use. By example, a dynamic spatial
panel methodology with fixed spatial an temporal effects (e.g.~Elhorst
2014, Cap. 4), but those models don't take account correlation between
errors. Therefore, a spatial SUR approach is more reasonable for our
purpose.

\hypertarget{analysis-and-results}{%
\section{Analysis and Results}\label{analysis-and-results}}

@@

We begin with the exploratory analysis of the data.

Figure \ref{fig:weekly-average-incidence-map} shows the geographical
variation in the incidence of the disease, as well as its temporal
progression in weekly averages. Our analysis begins on March 13, which
is the first date when every province had reported at least one case of
COVID-19. It can be seen that the highest incidence at this early date
was in the provice of Álava, in the North of Spain. While not the most
populous province, with a population of only 331,549, Álava has the
highest GDP per capita of all provinces. Vitoria, its main city, is the
capital of the Basque country and has been the focus of efforts to
develop a ``Global Basque City'' (Meijers et al., 2008), along with San
Sebastian and Bilbao. The other early focus of the pandemic in Spain was
Madrid, which is the most populous province in the country and has the
second highest GDP per capita after Álava. The early outbreaks in these
two provinces can be traced throughout the progression of the pandemic
over time, although by the end of the period under consideration, other
provinces had matched and even surpased their incidence rates, including
Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete
to the south.

@@

To visualize the distribution of temperature and humidity we aggregate
the provinces by Autonomous Community. In Figure
\ref{fig:descriptives-temperature} the communities have been sorted by
latitude, so that Principado de Asturias is the northernmost community,
and Canarias the southernmost. There is a relatively wide range of
values both within and between provinces over the 30 day period
examined. The top panel of the figure shows the distribution of mean
temperature. The lowest mean temperature for a community on any given
day is 2.8C, and the highest is 22.4C for a range of approximately 20
degrees. Likewise, there is a great deal of variability in humidity, as
seen in the bottom panel of the figure, where the lowest mean humidity
for any community is 48.3 and the highest is 99.6. The actual values for
the provinces display somewhat more variability even.

@@

@@

Figure \ref{fig:daily-correlations} shows the distribution of daily
correlations of the independent variables with incidence of COVID-19,
after log-transforming all variables. It can be seen there that the
correlation of GDPpc and temperature (in its three definitions) have the
strongest positive and negative correlations with incidence,
respectively. Percentage of older adults displays somewhat weaker
negative correlations with incidence, as does density. It can be seen
that the humidity variable, in its three forms, tends to be possitively
correlated with incidence of COVID-19.

@@

The goodness of fit of the three systems of equations is shown in Figure
\ref{fig:goodness-of-fit}.

@@

Summary of best model.

@@

\caption{\label{tab:summary-best-model}\label{tab:summary-best-model}Summary of estimation results best model (lagged 11-day moving average)}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrrrrrrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Estimates} & \multicolumn{3}{c}{Significance} & \multicolumn{1}{c}{ } \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7}
Variable & Min & Mean & Max & p > 0.10 & 0.10 <= p < 0.05 & p <= 0.05 & Note\\
\midrule
\rowcolor{gray!6}  Intercept & 7.370 & 10.037 & 12.968 & 0 & 0 & 30 & Non-constrained\\
log(GDPpc) & 0.620 & 0.620 & 0.620 & 0 & 0 & 1 & Constrained\\
\rowcolor{gray!6}  log(Older) & -0.737 & -0.737 & -0.737 & 0 & 0 & 1 & Constrained\\
log(Density) & -0.220 & -0.097 & 0.153 & 19 & 1 & 10 & Non-constrained\\
\rowcolor{gray!6}  Transit & 0.314 & 0.512 & 0.583 & 10 & 9 & 11 & Non-constrained\\
\addlinespace
log(Humidity) & -1.434 & -0.534 & -0.031 & 10 & 1 & 19 & Non-constrained\\
\rowcolor{gray!6}  log(Temperature) & -2.014 & -1.406 & -0.929 & 0 & 0 & 30 & Non-constrained\\
log(Sunshine) & -0.258 & 0.097 & 0.206 & 7 & 2 & 21 & Non-constrained\\
\bottomrule
\multicolumn{8}{l}{\textit{Note: }}\\
\multicolumn{8}{l}{Significance: This is the number of coefficients with p-values as indicated}\\
\multicolumn{8}{l}{Non-constrained: coefficient was allowed to vary across equations}\\
\multicolumn{8}{l}{Constrained: coefficient as constant across equations}\\
\end{tabular}}
\end{table}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Figure \ref{fig:delta-time} shows the temporal evolution of the spatial
autocorrelation coefficient (\(\lambda\)).

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/delta-time-1.pdf}
\caption{\label{fig:delta-time}Temporal evolution of the spatial
autocorrelation coefficient; dots are the point estimates and vertical
lines are 95\% confidence intervals. In yellow is the period after the
declaration of the state of emergency, and in orange is the period when
only essential activities were allowed.}
\end{figure}

Figure \ref{fig:beta-intercept-time} shows the temporal evolution of the
intercept.

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-intercept-time-1.pdf}
\caption{\label{fig:beta-intercept-time}Temporal evolution of intercept;
dots are the point estimates and vertical lines are 95\% confidence
intervals. In yellow is the period after the declaration of the state of
emergency, and in orange is the period when only essential activities
were allowed.}
\end{figure}

Figure \ref{fig:beta-density-time} shows the temporal evolution of the
coefficient for \(\log(Density)\).

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-density-time-1.pdf}
\caption{\label{fig:beta-density-time}Temporal evolution of coefficient
for log(Density); dots are the point estimates and vertical lines are
95\% confidence intervals. In yellow is the period after the declaration
of the state of emergency, and in orange is the period when only
essential activities were allowed.}
\end{figure}

Figure \ref{fig:beta-transit-time} shows the temporal evolution of the
coefficient for \(Transit\).

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-transit-time-1.pdf}
\caption{\label{fig:beta-transit-time}Temporal evolution of coefficient
for Transit; dots are the point estimates and vertical lines are 95\%
confidence intervals. In yellow is the period after the declaration of
the state of emergency, and in orange is the period when only essential
activities were allowed.}
\end{figure}

Figure \ref{fig:beta-humidity-time} shows the temporal evolution of the
coefficient for \(\log(Humidity)\).

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-humidity-time-1.pdf}
\caption{\label{fig:beta-humidity-time}Temporal evolution of coefficient
for log(Humidity); dots are the point estimates and vertical lines are
95\% confidence intervals. In yellow is the period after the declaration
of the state of emergency, and in orange is the period when only
essential activities were allowed.}
\end{figure}

Figure \ref{fig:beta-temperature-time} shows the temporal evolution of
the coefficient for \(\log(Temperature)\).

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-temperature-time-1.pdf}
\caption{\label{fig:beta-temperature-time}Temporal evolution of
coefficient for log(Temperature); dots are the point estimates and
vertical lines are 95\% confidence intervals. In yellow is the period
after the declaration of the state of emergency, and in orange is the
period when only essential activities were allowed.}
\end{figure}

Figure \ref{fig:beta-sunshine-time} shows the temporal evolution of the
coefficient for \(\log(Sunshine + 0.1)\).

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-sunshine-time-1.pdf}
\caption{\label{fig:beta-sunshine-time}Temporal evolution of coefficient
for log(Sunshine + 0.1); dots are the point estimates and vertical lines
are 95\% confidence intervals. In yellow is the period after the
declaration of the state of emergency, and in orange is the period when
only essential activities were allowed.}
\end{figure}

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{A spatio-temporal analysis of the environmental correlates of COVID-19
incidence in Spain}
    \author[McMaster University]{Antonio Paez\corref{1}}
   \ead{paezha@mcmaster.ca} 
    \author[Universidad Politecnica de Cartagena]{Fernando A. Lopez}
   \ead{fernando.lopez@upct.es} 
    \author[Departamento de Economia]{Tatiane Menezes}
   \ead{tatiane.menezes@ufpe.br} 
    \author[Nucleo de Pesquisa]{Renata Cavalcanti}
   \ead{renata.vcsantos@gmail.com} 
    \author[Nucleo de Pesquisa]{Maira Galdino da Rocha Pitta}
   \ead{mgrpitta@ufpe.br} 
      \address[McMaster University]{School of Geography and Earth Sciences, McMaster University, 1281 Main
St W, Hamilton, ON, L8S 4K1, Canada}
    \address[Universidad Politecnica de Cartagena]{Departamento de Metodos Cuantitativos, Ciencias Juridicas, y Lenguas
Modernas, Universidad Politecnica de Cartagena, Calle Real Numero 3,
30201, Cartagena, Murcia, Spain}
    \address[Departamento de Economia]{Departamento de Economia, Universidade Federal de Pernambuco, Av dos
Economistas, s/n - Cidade Universitária, Recife - PE, 50670-901, Brasil}
    \address[Nucleo de Pesquisa]{Núcleo de Pesquisa em Inovação Terapêutica NUPIT / UFPE, Av.
Prof.~Moraes Rego, 1235 - Cidade Universitária, Recife, PE, CEP
50670-901, Brazil}
      \cortext[1]{Corresponding Author}

@@

\hypertarget{context-and-data}{%
\section{Context and Data}\label{context-and-data}}

@@

The first variable that we consider is GDP per capita. Much has been
said about globalization and the spread of infectious
disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p. A18)}.
The growth in global connections has presented a challenge to spatial
approaches in the initial stages of disease management, when the cause
of a disease may still be unclear, but the plane has already departed
(Zhou and Coleman, 2016). In reference to the earlier outbreak of SARS,
van Wagner (Van Wagner, 2008) chronicles how Toronto's status as a
global city turned out to be a vulnerability in this respect. In our
case, we think of GDP per capita as a marker of a region's relative
position in a network of global cities, and its potential to be further
ahead in the trajectory of the pandemic. Furthermore, wealthier regions
also tend to concentrate more activities that produce non-traded goods,
including building and construction (Hallet, 2002). Therefore, it is
possible that wealthier regions remain relatively more active even
during a lockdown. On the other hand, it is also possible that less
wealthy regions have a higher proportion of workers in manual
occupations that cannot telework, and therefore have more difficulties
complying with shelter-in-place orders.

The percentaje of older adults (over 65) is the second variable that we
consider as a control. Early evidence regarding COVID-19 suggests that
the case rate mortality is higher at older ages (e.g.~The Novel
Coronavirus Pneumonia Emergency Response Epidemiology Team, 2020).
However, it is not clear that a relatively large population of older
adults necessarily translates into higher transmission of the infection.
The tool of choice in containing the spread of the disease has been
social distancing. In this respect, the evidence from the field of
transportation is that older adults tend to travel less frequently, for
shorter distances, and have higher rates of immobility (e.g., Roorda et
al., 2010; Morency et al., 2011; Sikder and Pinjari, 2012). In other
words, many older adults are, whether by preference or otherwise,
already in a form of social isolation. Social distancing during the
pandemic may actually reinforce that condition for them, as suggested by
the analysis of age-structured social contact in India, China, and Italy
of Sing and Adhikari (2020). Accordingly, our expectation is that
provinces with higher percentages of older adults will tend to have
similarly low levels of social contact, particularly since the
age-structured matrix of social contact in Spain is similar to Italy
(Prem et al., 2017).

@@

The last control variable is the presence of mass transit systems in a
province. Every province in Spain offers some form of public
transportation, however only five provinces have higher order systems of
mass mobility (e.g.~metro or subway), namely Barcelona, Madrid, Sevilla,
Valencia, and Bizkaia. Public transportation has been hypothesized to
relate to the spread of contagious disease by some researchers using
agent-based approaches and simulation (e.g., Perez and Dragicevic, 2009;
Wang et al., 2010), and while we find scant evidence of a link in the
literature, the idea is intuitively appealing. After all, unlike the
isolation that a car offers to travellers, most mass transit system are
cauldrons of social contact.

@@

\hypertarget{methods-the-spatial-sur-model}{%
\section{Methods: the Spatial SUR
Model}\label{methods-the-spatial-sur-model}}

@@

The SUR-SLM model can be estimated using maximum likelihood (López et
al. (2014)) or instrumental variables (Mínguez et al. (2019)).

\hypertarget{analysis}{%
\section{Analysis}\label{analysis}}

@@

Figure \ref{fig:weekly-average-incidence-map} shows the geographical
variation in the incidence of COVID-19 in Spain, as well as the temporal
progression of the disease in weekly averages. Our analysis begins on
March 13, which is the first date when every province had reported at
least one case of COVID-19. It can be seen that the highest incidence at
this early date was in the provice of Álava, in the North of Spain.
While not the most populous province, with a population of only 331,549,
Álava has the highest GDP per capita of all provinces. Vitoria, its main
city, is the capital of the Basque Country and has been the focus of
efforts, along with San Sebastian and Bilbao, to develop a ``Global
Basque City'' (Meijers et al., 2008). The other early focus of the
pandemic in Spain was Madrid, which is the most populous province in the
country and has the second highest GDP per capita after Álava. The early
outbreaks in these two provinces can be traced throughout the
progression of the pandemic over time, although by the end of the period
under consideration, other provinces had matched and even surpased their
incidence rates, including Segovia and Soria to the north of Madrid, and
Ciudad Real and Albacete to the south.

@@

Figure \ref{fig:descriptives-temperature} shows the distribution of the
environmental variables in Spain. For ease of visualization we aggregate
the provinces by Autonomous Community. Each box-and-whisker in the
figure represents the distribution of the variable for a community over
the 30-day period. In the plot, the communities have been sorted by
latitude, so that Principado de Asturias is the northernmost community,
and Canarias the southernmost. As seen in the figure, there is a
relatively wide range of values both within and between provinces over
the 30-day period examined. The top panel of the figure shows the
distribution of mean temperatures. The lowest mean temperature for a
community on any given day was 2.8C, and the highest 22.4C, for a range
of approximately 20 degrees. Likewise, there is a great deal of
variability in humidity, as seen in the middle panel of the figure,
where the lowest mean humidity for any community is 48.3 and the highest
is 99.6. Finally, the bottom panel displays mean daily hours of sunshine
in the community. This variable displays the most variability within
communities over time, but remains relatively constant across
communities. It is important to note that these are summaries by
community, and the actual values of these variables for the provinces
display somewhat more variability.

@@

Figure \ref{fig:map-controls} includes three maps that display the
spatial variation of our control variables, namely GDP per capita,
percentage of older adults in province, population density, and presence
of mass transit systems. As seen there, GDP per capita is higher in
Madrid and the northeast part of the country, mainly in Pais Vasco and
Cataluña. Percentage of older adults is somewhat more checkered, with
high values in Madrid and other provinces in the center-west part of the
country, but also in some provinces in the north. Outside of provinces
with major cities (e.g., Madrid; Bizkaia and Gipuzkoa in Pais Vasco;
Pontevedra in Galicia), population density tends to be higher in
provinces along the Mediterranean coast. The final panel in the figure
shows the five provinces in the country that have higher order mass
transit systems (e.g., metro).

@@

Figure \ref{fig:daily-correlations} shows the distribution of daily
simple correlations of incidence of COVID-19 with the independent
variables (with the exception of Transit, which is a categorical
variable). These correlations are calculated after log-transforming all
variables. As previously discussed, the environmental variables have a
temporal lag and were calculated different time windows.

It can be seen in the figure that temperature (in its three forms) has
the highest simple correlation with incidence of COVID-19. After
temperature, GDP per capita has the highest positive correlation with
the dependent variable. The distribution of these correlations is also
quite tight over the 30-day period of the study. Hours of sunshine tends
to have a moderately high correlation with incidence of COVID-19,
however the distribution of these correlations is more spread, and in
some cases strays into negative values. A similar thing happens with
humidity, which also tends to display more day to day variation in the
correlation with the dependent variable. The percentage of older adults
shows a relatively tight distribution of day-to-day correlations, and is
negative. Population density, in contrast, tends to be negative, but is
relatively spread, and on some days, the simple correlation between
density and incidence of COVID-19 is weakly positive.

@@

Correlation analysis in the preceding section provides some insights
about the potential associations between incidence of COVID-19 and the
various environmental and control variables. In this section we estimate
three spatial SUR models to test the differences between the various
temporal lags and weighting schemes for the environmental variables.
Accordingly, we define three models: Model 1, which is estimated using
the lagged 8-day averages of the environmental variables (\emph{lag8});
Model 2, which is estimated using the lagged 11-day averages of the
environmental variables (\emph{lag11}); and finally, Model 3, which is
estimated using the lagged 11-day \emph{weighted} averages of the
environmental variables (\emph{lag11w}).

To implement the SUR approach, we must define a matrix of spatial
weights \(W\). In this case, we consider neighborhoods based on
adjacency, based on the well-known rook criterion (two provinces are
adjacent if they share a boundary, but not if they touch only at a
vertex). We modify this criterion in two cases. First, we make an
allowance for adjacency between the two islands in the Autonomous
Community of Canarias in the Pacific (Las Palmas and Santa Cruz de
Tenerife), which we assume are adjacent between each other. And
secondly, we assume that Islas Baleares in the Mediterranean are
adjacent to three provinces in Pais Catalans (i.e., Barcelona,
Tarragona, and Castello). After matrix \(W\) has been specified,
estimation of the model can proceed as usual.

For estimation, we log-transform the dependent and quantitative
independent variables. The only variable that is not transformed is the
categorical variable for transity systems. A log-log transformation is
appropriate to capture non-linear relationships between variables and
provides a straightforward interpretation of the coefficients as
percentage change. Furthermore, we introduce restrictions so that the
coefficients of two of our control variables are constant over time,
namely GDP per capita and percentage of older adults. We do not see an
\emph{a priori} reason to let those two variables vary across equations,
and the correlation analysis in Figure \ref{fig:daily-correlations} also
suggest little temporal variation. In contrast, we let the spatial
autocorrelation parameter, as well as the parameters of the rest of the
independent variables (including the constant) to vary over
time\footnote{We conducted sensitivity analysis letting all parameters vary over time, and while the results are qualitatively similar, the resulting models are less parsimonious. These results are available in the source R markdown document.}.

After estimation, we compare the goodness of fit of the three SUR
models. Figure \ref{fig:goodness-of-fit} shows the equation-level
coefficient of determination \(R^2\), one for each time period/day. As
well, the overall coefficient of determination for the system is
reported for each model \(\text{pooled}-R^2\). The general trend is
identical for the three models, with the goodness-of-fit improving over
time and plateuing around a value of \(R^2\) of 0.6. Model 1
(\emph{lag8}) performs somewhat better in the first few equations/days,
when the goodness-of-fit is relatively poor, and then again in the last
few equations/days. Model 3 (\emph{lag11w}), in contrast, does not
perform well towards the end of the study period. The most balanced
model in terms of equation-level goodness-of-fit appears to be Model 1
(\emph{lag11}), and this impression is further supported with a slightly
higher value of the \(\text{pooled}-R^2\). This analysis is in line with
the incubation period reported by Lauer et al.~(2020), although not the
use of a weighted average. For the remainder of the paper, we will adopt
Model 2 (\emph{lag11}) as our best model. In the following section we
discuss the results of the analysis in more depth.

@@

\hypertarget{results-and-discussion}{%
\section{Results and Discussion}\label{results-and-discussion}}

Table \ref{tab:summary-best-model} presents a summary.

@@

\caption{\label{tab:summary-best-model}\label{tab:summary-best-model}Summary of estimation results of best model (lag11: lagged 11-day moving average)}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lrrrrrrl}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Estimates} & \multicolumn{3}{c}{Significance} & \multicolumn{1}{c}{ } \\
\cmidrule(l{3pt}r{3pt}){2-4} \cmidrule(l{3pt}r{3pt}){5-7}
Variable & Min & Mean & Max & p > 0.10 & 0.10 <= p < 0.05 & p <= 0.05 & Note\\
\midrule
\rowcolor{gray!6}  Intercept & 7.370 & 10.037 & 12.968 & 0 & 0 & 30 & Non-constrained\\
log(GDPpc) & 0.620 & 0.620 & 0.620 & 0 & 0 & 1 & Constrained\\
\rowcolor{gray!6}  log(Older) & -0.737 & -0.737 & -0.737 & 0 & 0 & 1 & Constrained\\
log(Density) & -0.220 & -0.097 & 0.153 & 19 & 0 & 11 & Non-constrained\\
\rowcolor{gray!6}  Transit & 0.314 & 0.512 & 0.583 & 8 & 8 & 14 & Non-constrained\\
\addlinespace
log(Humidity) & -1.434 & -0.534 & -0.031 & 10 & 1 & 19 & Non-constrained\\
\rowcolor{gray!6}  log(Temperature) & -2.014 & -1.406 & -0.929 & 0 & 0 & 30 & Non-constrained\\
log(Sunshine) & -0.258 & 0.097 & 0.206 & 7 & 2 & 21 & Non-constrained\\
\rowcolor{gray!6}  Spatially lagged y (rho) & 0.015 & 0.142 & 0.528 & 12 & 3 & 15 & Non-constrained\\
\bottomrule
\multicolumn{8}{l}{\textit{Note: }}\\
\multicolumn{8}{l}{Significance: This is the number of coefficients with p-values as indicated}\\
\multicolumn{8}{l}{Non-constrained: coefficient was allowed to vary across equations}\\
\multicolumn{8}{l}{Constrained: coefficient as constant across equations}\\
\end{tabular}}
\end{table}

Figure \ref{fig:results-time-1} shows the temporal evolution of the
spatial autocorrelation coefficient (\(\rho\)) and the intercept of the
model.

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/results-time-1-1.pdf}
\caption{\label{fig:results-time-1}Temporal evolution of the spatial
autocorrelation coefficient (rho) and the intercept of the model; dots
are the point estimates and vertical lines are 95\% confidence
intervals. In yellow is the period after the declaration of the state of
emergency, and in orange is the period when only essential activities
were allowed.}
\end{figure}

Figure \ref{fig:beta-controls-time} shows the temporal evolution of the
coefficients for the two control variables that were not fixed over
time, i.e., \(\log(Density)\) and \(Transit\).

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-controls-time-1.pdf}
\caption{\label{fig:beta-controls-time}Temporal evolution of coefficient
for the control variables; dots are the point estimates and vertical
lines are 95\% confidence intervals. In yellow is the period after the
declaration of the state of emergency, and in orange is the period when
only essential activities were allowed.}
\end{figure}

Figure \ref{fig:beta-environmental-time} shows the temporal evolution of
the coefficient for the three environmental variables.

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/beta-environmenta-time-1.pdf}
\caption{\label{fig:beta-environmental-time}Temporal evolution of
coefficient for the environmental variables; dots are the point
estimates and vertical lines are 95\% confidence intervals. In yellow is
the period after the declaration of the state of emergency, and in
orange is the period when only essential activities were allowed.}
\end{figure}
